### 2020-07-17 paper.tex
clean up automated involutive mccm algorihtm box

\textbf{Relationship to representation of~\citet{green1995reversible}.}
\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a Eucliean space of parameters.
Dictionaries possesses substantially more structure:
Parameters have keys, so it is possible to express that parameters are shared across different structures.
Instead of a monolithic `model', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys.
The additional structure of dictionaries enables the automation technique Section~\ref{sec:automating}.

@@

Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is an integral over $\mathbb{R}^{2k}$.

@@

Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}$, so that $Z = X \times Y \subseteq \mathcal{D} \times \mathcal{D}$.
%And suppose that there exists an involution $f : Z \to Z$ such that (i) 
Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant (so that each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries).
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Then, the acceptance probability simplifies to:
\begin{equation}
\alpha(x, y) = \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

@@

\begin{algorithm*}[h]
\begin{algorithmic}
\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
    \State $(y, \log w_2) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$ and compute log-density $\log q_x(y)$}
    \State $(x', y', \log D) \gets \textproc{run-involution}(\mathcal{F}, (x, y))$ \Comment{Compute $(x', y') = f(x, y)$, log Radon-Nikodym deriv. $\log D$}
    \State $\log w_1 \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density $\log p(x)$}
    \State $\log w_3 \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density $\log p(x')$}
    \State $\log w_4 \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density $\log q_{x'}(y')$}
    \State $\alpha \gets \min\{1, \exp(\log w_3 - \log w_1 + \log w_4 - \log w_2 + \log J)\}$ \Comment{Compute acceptance probability}
    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure\\

\begin{minipage}[t]{0.5\linewidth}
\Procedure{trace-and-score}{$\mathcal{P}$}
   \State $\log w \gets 0$
   \State $x \gets \{\}$
   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
        \State \hspace{5mm} 1. set $v \sim d$
        \State \hspace{5mm} 2. set $x[a] \gets v$
        \State \hspace{5mm} 3. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
        \State \hspace{5mm} 4. evaluate to $v$)
    \END
   \State \Return $(x, \log w)$
\EndProcedure
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\Procedure{score}{$\mathcal{P}$, $x$}
   \State $\log w \gets 0$
   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
        \State \hspace{5mm} 1. set $v \gets x[a]$
        \State \hspace{5mm} 2. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
        \State \hspace{5mm} 3. evaluate to $v$)
    \END
   \State \Return $\log w$
\EndProcedure
\end{minipage}\\

\Procedure{run-involution}{$\mathcal{F}$, $(x, y)$}
% TODO would be much simpler to describe if we didn't have to separate out x and y..
    \LineComment{Run the involution, and keep track of continuous addresses}
    \State $x' \gets \{\}$ \Comment{Initialize empty output dictionary}
    \execute[$\mathcal{F}$], but with
        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $x'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $x[a]$)
        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $x'[a_2] \gets x[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
    \END
    \LineComment{Use automatic differentiation to compute Jacobian, skipping copied addresses}
    \For{$i$ in $1, \ldots, |W \setminus C|$}
        \State $J[i,:] \gets \nabla_{x_{R \setminus C}} \left( h_{e(x, y)}[(W \setminus C)_i] \right)$ \Comment{Gradient of $x'[(W \setminus C)_i]$ w.r.t. non-copied continuous inputs}
    \EndFor
    \State \Return $(x', \log |\det(J)|)$
\EndProcedure

\end{algorithmic}
\caption{Automated Involutive MCMC}
\label{alg:auto-involutive-mcmc}
\end{algorithm*}

@@

In Algorithm~\ref{alg:involutive-mcmc} the involution $f : Z \to Z$ maps a pair $(x, y) \in Z$ to a new pair $(x', y') \in Z$.
When the densities $p$ and $q$ are defined with probabilistic programs, each element of these pairs is a trace.
Recall that a trace is mathematically, a finite dictionary that maps addresses to values.
Therefore, the involution is a map from pairs of traces to pairs of traces.
%To simplify the notation, we can represent a pair of traces $(x, y)$ as a single trace $z$, in which each address in $x$ and $x'$ is prefixed with some unique identifier (e.g. \texttt{model}) and each address in $y$ is prefixed with some unique identifier $y'$ (e.g. \texttt{auxiliary}).
Then, $Z$ is a set of such trace pairs, and the involution $f$ is a function from $Z$ onto itself that is its own invverse.

>>>>

\textbf{Relationship to representation of~\citet{green1995reversible}.}
\citet{green1995reversible} uses a state space that is the countable union of `models', where for each model there is a vector of real-valued parameters.
Dictionaries possesses substantially more structure:
Instead of monolithic `models', dictionaries use a more elaborate discrete state that includes the set of keys and the assignment to the discrete keys.
Also, continuous keys play the role of real-valued parameters, so it is possible to express that a given real-valued parameter is shared between models.
The additional structure of dictionaries enables the automation techniques in Section~\ref{sec:automating}.

@@

Consider the infinite univariate mixture model, and the conditional density given observed data $(B, \mathbf{b}) := \{\mathtt{y}_1 \mapsto y_1, \ldots, \mathtt{y}_{\mathtt{n}} \mapsto y_{\mathtt{n}}\}$.
The conditional density $p(d | b)$ is nonzero only if $A_d = \{\texttt{k}\} \cup \{\texttt{m}_i \mid 1 \leq i \leq k\} \cup \{\texttt{s}_i \mid 1 \leq i \leq k\}$ for some $k$ ($d$ does not contain y-values), and the denominator in the definition of $p(d | b)$ simplifies to the familiar sum of marginal likelihoods over all $k$, where each marginal likelihood is a Riemann integral over $\mathbb{R}^{2k}$.

@@

Suppose that the model distribution and auxiliary distributions are probability distributions on dictionaries, with densities $p$ and $q_x$.
Then, $X$ and $Y$ are both sets of dictionaries, and the joint space $Z$ is a set of pairs $(x, y)$ of dictionaries with keys $\mathcal{A}_P$ and $\mathcal{A}_Q$ respectively, so that $Z = X \times Y \subseteq \mathcal{D}_P \times \mathcal{D}_Q$ where $\mathcal{D}_P$ is the set of dictionaries on keys taken from $\mathcal{A}_P$ and similarly for $\mathcal{D}_Q$.
% TODO: this is only needed for Algorithm 2, and it is relevant for the 
To simplify the notation, and without loss of generality, we will assume that $\mathcal{A}_P$ and $\mathcal{A}_Q$ are disjoint\footnote{If $\mathcal{A}_P$ and $\mathcal{A}_Q$ are not disjoint then, they can be made so by adding a different prefix to each address.}, and we define $Z := \{ x \oplus y : p(x)q_x(y) > 0\} \subseteq \mathcal{D}$.
where $\mathcal{D}$ is the set of dictionaries on keys from $\mathcal{A}_P \cup \mathcal{A}_Q$
(recall $x \oplus y$ denotes the dictionary resulting from merging dictionaries $x$ and $y$ with disjoint keys).

Suppose there is a countable partition of $Z$ into $\{Z_e : e \in E\}$ such that across each $Z_e$ the sets of keys $A_x$ and $A_y$ and the values $x_a$ and $y_a$ for all discrete keys $a$ are both constant.
Then each set $Z_e$ is isomorphic to a Euclidean space of assignments to the continuous keys in the two dictionaries.
Suppose there is an involution $g : E \to E$ between elements of the partition (a function with $g = g^{-1}$), and a family of continuously differentiable bijections $h_{e} : Z_e \to Z_{g(e)}$ for each $e$ with $h_{e} = h_{g(e)}^{-1}$.
Let $e(z) \in E$ denote the element of the partition for an alement $z \in Z$.
Then, $f : Z \to Z$ given by $f(z) := h_{e(z)}(z)$ is an involution:
\begin{equation*}
f(f(z)) = h_{e(f(z))}(h_{e(z)}(z)) = h_{g(e(z))}(h_{e(z)}(z)) = z
\end{equation*}
Let $|J h_e|(z)$ denote the absolute value of the Jacobian (determinant) of $h_e$, evaluated at $z$.
Then, the acceptance probability in Algorithm~\ref{alg:involutive-mcmc} simplifies to:
\begin{equation}
\frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \cdot |J h_e|(z)
\end{equation}
One example of a valid partition of $Z$ is given by equivalence classes of the following equivalence relation:
\[
z_1 \sim z_2 \iff (A_{z_1} = A_{z_2}) \land (z_1[a] = z_2[a] \; \forall a \in A_{z_1} \cap \mathcal{I})
\]
(dictionaries are equivalent if they contain the same keys and they agree on the value of all discrete keys).
Section~\ref{sec:radon-nikodym-special-case} of the appendix for details.

@@

\begin{algorithm*}[h]
\begin{algorithmic}
\Require probabilistic programs $\mathcal{P}$ and $\mathcal{Q}$; differentiable program $\mathcal{F}$; initial state $x$
\vspace{1mm}
\Procedure{auto-involutive-mcmc}{$\mathcal{P}$, $\mathcal{Q}$, $\mathcal{F}$, $x$}
    \State $(y, \log q_x(y)) \gets \textproc{trace-and-score}(\mathcal{Q}_x)$ \Comment{Sample $y \sim q_x(\cdot)$; compute log-density via probabilistic program}
    \State $(x' \oplus y', \log D) \gets \textproc{run-involution}(\mathcal{F}, x \oplus y)$ \Comment{{\small Compute $(x' \oplus y') = f(x \oplus y)$ and log(Radon-Nikodym derivative)}}
    \State $\log p(x) \gets \textproc{score}(\mathcal{P}, x)$ \Comment{Compute log-density via probabilistic program}
    \State $\log p(x') \gets \textproc{score}(\mathcal{P}, x')$ \Comment{Compute log-density via probabilistic program}
    \State $\log q_{x'}(y') \gets \textproc{score}(\mathcal{Q}_{x'}, y')$ \Comment{Compute log-density via probabilistic program}
    \State $\alpha \gets \min\{1, \exp(\log p(x') - \log p(x) + \log q_{x'}(y') - \log q_x(y) + \log D)\}$ \Comment{Compute acceptance probability}
    \State \textbf{with probability} \,$\alpha$ \,\Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure\\
\vspace{1mm}
\begin{minipage}[t]{0.5\linewidth}
\Procedure{trace-and-score}{$\mathcal{P}$}
   \State $\log w \gets 0$
   \State $x \gets \{\}$
   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
        \State \hspace{5mm} 1. set $v \sim d$
        \State \hspace{5mm} 2. set $x[a] \gets v$
        \State \hspace{5mm} 3. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
        \State \hspace{5mm} 4. evaluate to $v$)
    \END
   \State \Return $(x, \log w)$
\EndProcedure
\end{minipage}%
\begin{minipage}[t]{0.5\linewidth}
\Procedure{score}{$\mathcal{P}$, $x$}
   \State $\log w \gets 0$
   \execute[$\mathcal{P}$], but with $\mathtt{"a \sim d"} \equiv$ (
        \State \hspace{5mm} 1. set $v \gets x[a]$
        \State \hspace{5mm} 2. set $\log w \gets \log w + \textproc{logpdf}(d, v)$
        \State \hspace{5mm} 3. evaluate to $v$)
    \END
   \State \Return $\log w$
\EndProcedure
\end{minipage}\\
\vspace{1mm}
\Procedure{run-involution}{$\mathcal{F}$, $z$}
% TODO would be much simpler to describe if we didn't have to separate out x and y..
    \LineComment{Run the involution, and keep track of copied, read, and written continuous keys}
    \State $z' \gets \{\}$ \Comment{Initialize empty output dictionary}
    \execute[$\mathcal{F}$], but with
        \State $\mathtt{"write\_disc}(a, v)\mathtt{"} \equiv$ (set $z'[a] \gets v$)
        \State $\mathtt{"read\_disc}(a)\mathtt{"} \equiv$ (evaluate to $z[a]$)
        \State $\mathtt{"write\_cont}(a, v)\mathtt{"} \equiv$ (1. set $z'[a] \gets v$ and 2. set $W \gets W \cup \{a\}$)
        \State $\mathtt{"read\_cont}(a)\mathtt{"} \equiv$ (1. set $R \gets R \cup \{a\}$ and 2. evaluate to $z[a]$)
        \State $\mathtt{"copy}(a_1, a_2)\mathtt{"} \equiv$ (1. set $z'[a_2] \gets z[a_1]$ and 2. set $C \gets C \cup \{a_2\}$)
    \END
    \LineComment{Use automatic differentiation of $\mathcal{F}$ to compute Jacobian, skipping copied addresses}
    \For{$i$ in $1, \ldots, |W \setminus C|$}
        \State $a_i \gets (W \setminus C)_i$ \Comment{Pick $i$th key in $(W \setminus C)$; the order does not matter}
        \State $J[i,:] \gets \nabla_{\mathbf{z}_{R \setminus C}} \left( h_{e(z)}[a_i] \right)$ \Comment{{\small Gradient of $z'[a_i] = h_{e(z)}[a_i]$ wrt non-copied continuous inputs ($z_a$ for $a \in R \setminus C$)}}
    \EndFor
    \State \Return $(z', \log(|\det(J)|))$
\EndProcedure
\end{algorithmic}
\caption{Automated Involutive MCMC}
\label{alg:auto-involutive-mcmc}
\end{algorithm*}

@@

In Algorithm~\ref{alg:involutive-mcmc} the involution $f : Z \to Z$ maps a pair $(x, y) \in Z$ to a new pair $(x', y') \in Z$.
When the densities $p$ and $q$ are defined with probabilistic programs, each element of these pairs is a trace.
Recall that a trace is mathematically, a finite dictionary that maps addresses to values.
Therefore, the involution is a map from pairs of traces to pairs of traces.
Then, $Z$ is a set of such trace pairs, and the involution $f$ is a function from $Z$ onto itself that is its own invverse.
